---
title: "`SparseAssays` vs. `JointSparseAssays`"
author: "Peter Hickey"
date: "13 October 2015"
output: 
  html_document: 
    keep_md: yes
---

## Overview

I need to benchmark the `SparseAssays` class with the `JointSparseAssays` class using WGBS data for various sized m-tuples, with $NIL = 0$ and $NIL \geq 0$. I use the _ADS_ sample from the _Lister_ dataset.

# Installation

To install _SparseSummarizedExperiment_, we need to use the development version of Bioconductor. We use [_packrat_](https://cran.r-project.org/web/packages/packrat/index.html) to avoid polluting the general R library:

```{r packrat-setup}
if (!("packrat" %in% installed.packages())) {
  install.packages("packrat")
}
library(packrat)
packrat::init()
```

Now, install the development version of Bioconductor:

```{r bioc-setup}
source("http://bioconductor.org/biocLite.R")
useDevel()
```

Now, install _SparseSummarizedExperiment_:

```{r sse-setup}
install.packages("devtools")
devtools::install_github("PeteHaitch/SparseSummarizedExperiment")
library(SparseSummarizedExperiment)
```

Also use 

- [_data.table_](https://cran.r-project.org/web/packages/data.table/index.html) for reading in files
- [_pryr_](https://cran.r-project.org/web/packages/data.table/index.html) for inspecting R objects
- [_GenomicTuples_](http://bioconductor.org/packages/GenomicTuples/) for storing genomic co-ordinates

```{r data.table}
install.packages(c("data.table", "pryr"))
library(data.table)
library(pryr)
library(GenomicTuples)
```

# Convert `methtuple` output to a `MethPat` and `SparseMethPat` objects

## Read in some data

```{r ingest}
ADS_2 <- fread("DM_ADS.CG.2.tsv")
pryr::object_size(ADS_2)

ADS_2ac <- fread("DM_ADS.CG.2ac.tsv")
pryr::object_size(ADS_2ac)

ADS_4 <- fread("DM_ADS.CG.4.tsv")
pryr::object_size(ADS_4)
```

## `SparseMethPat` object

### Convert genomic co-ordinates into `MTuples` object

```{r dt2mt}
mt_ADS_2 <- MTuples(gtuples = GTuples(seqnames = ADS_2[, chr],
                                      tuples = as.matrix(ADS_2[, .(pos1, pos2)]),
                                      strand = ADS_2[, strand]),
                    methinfo = MethInfo("CG"))
mt_ADS_2ac <- MTuples(gtuples = GTuples(seqnames = ADS_2ac[, chr],
                                        tuples = as.matrix(ADS_2ac[, .(pos1, pos2)]),
                                        strand = ADS_2ac[, strand]),
                      methinfo = MethInfo("CG"))
mt_ADS_4 <- MTuples(gtuples = GTuples(seqnames = ADS_4[, chr],
                                      tuples = as.matrix(ADS_4[, .(pos1, pos2, pos3, pos4)]),
                                      strand = ADS_4[, strand]),
                    methinfo = MethInfo("CG"))
```

### Convert counts into `SparseAssays` object

```{r dt2sa}
sa_ADS_2 <- SparseAssays(
  SimpleList(counts = 
               SimpleList(ADS = SparseSummarizedExperiment:::.sparsify(ADS_2[, .(MM, MU, UM, UU)]))))
pryr::object_size(sa_ADS_2)

sa_ADS_2ac <- SparseAssays(
  SimpleList(counts = 
               SimpleList(ADS = SparseSummarizedExperiment:::.sparsify(ADS_2ac[, .(MM, MU, UM, UU)]))))
pryr::object_size(sa_ADS_2ac)

sa_ADS_4 <- SparseAssays(
  SimpleList(counts = 
               SimpleList(ADS = SparseSummarizedExperiment:::.sparsify(ADS_4[, .(MMMM, MMMU, MMUM, MMUU, MUMM, MUMU, MUUM, MUUU, UMMM, UMMU, UMUM, UMUU, UUMM, UUMU, UUUM, UUUU)]))))
pryr::object_size(sa_ADS_4)
```


### Convert counts into `JointSparseAssays` objects

__TODO__: The `JointSparseAssays` class is rather fiddly for which to define `[` and `[<-` methods. Working on this is a low priority, but I will try to make methods for `RangedSummarizedExperiment` object agnostic to whether the `SparseAssays` slot is a `SparseAssays` or `JointSparseAssays` object.

### Create the `SparseMethPet` object

```{r dt2smp}
smp_ADS_2 <- SparseMethPat(sparseAssays = sa_ADS_2, 
                           rowTuples = mt_ADS_2,
                           colData = DataFrame(row.names = "ADS"))
pryr::object_size(smp_ADS_2)
smp_ADS_2ac <- SparseMethPat(sparseAssays = sa_ADS_2ac, 
                             rowTuples = mt_ADS_2ac,
                             colData = DataFrame(row.names = "ADS"))
pryr::object_size(smp_ADS_2ac)
smp_ADS_4 <- SparseMethPat(sparseAssays = sa_ADS_4, 
                           rowTuples = mt_ADS_4,
                           colData = DataFrame(row.names = "ADS"))
pryr::object_size(smp_ADS_4)
```
